---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

```{r}
library(tidyverse)
library(lubridate)
```

```{r}
activity <- read_csv("./activity.zip", col_names = TRUE)
```

## What is mean total number of steps taken per day?

```{r}
dailysteps <- activity %>% group_by(date) %>% summarize(steps = sum(steps))
dailysteps
```
```{r}
hist(dailysteps$steps, breaks = 10)
```

```{r}
mean(dailysteps$steps, na.rm = T)
median(dailysteps$steps, na.rm = T)
```

## What is the average daily activity pattern?

```{r}
intsteps <- activity %>% group_by(interval) %>% summarise(mean_steps = mean(steps, na.rm = T))
intsteps
```

```{r}
with(intsteps, plot(interval, mean_steps, type = "l"))
```

```{r}
intsteps$interval[which.max(intsteps$mean_steps)]
```

## Imputing missing values

How many rows in the data set have NAs?

```{r}
sum(is.na(activity))
```

Let's isolate those into a separate tibble and then join them with the tibble containing average steps by interval.

```{r}
actNA <- activity[is.na(activity$steps),]
NAjoin <- left_join(actNA, intsteps)
head(NAjoin)
```

Now let's assign the mean steps per interval to their corresponding NA values in a new data set.

```{r}
activitynoNA <- activity
activitynoNA$steps[is.na(activitynoNA$steps)] <- NAjoin$mean_steps
mean(is.na(activitynoNA$steps))
head(activitynoNA)
```

```{r}
dailystepsnoNA <- activitynoNA %>% group_by(date) %>% summarise(total_steps = sum(steps))
head(dailystepsnoNA, n = 10)
```

Here's a histogram of the daily steps with the imputed NA values.

```{r}
hist(dailystepsnoNA$total_steps, breaks = 10)
```

Let's plot the daily steps with NAs and with NA values imputed side by side to see the difference.

```{r}
par(mfrow = c(1,2))
hist(dailysteps$steps, breaks = 10, main = "Total Daily Steps w/ NAs", xlab = "Total Steps", ylim = c(0,25))
hist(dailystepsnoNA$total_steps, breaks = 10, main = "Total Daily Steps, Imputed NAs", xlab = "Total Steps", ylim = c(0,25))
```

Now let's examine the means and medians of the imputed and original data sets.

```{r}
DailySum <- matrix(list("Imputed", "NAs", mean(dailystepsnoNA$total_steps), mean(dailysteps$steps, na.rm = T), median(dailystepsnoNA$total_steps), median(dailysteps$steps, na.rm = T)), 2, 3)
colnames(DailySum) <- c("Cat", "Mean", "Median")
DailySum
```

It seems that imputing the NA values from the mean by interval changed the median to equal the mean value, but had no effect on the mean.

## Are there differences in activity patterns between weekdays and weekends?

First let's create a new factor variable indicating whether the observation occurred on a weekday or weekend.

```{r}
activitynoNA <- activitynoNA %>% mutate(DayClass = as.factor(ifelse(wday(ymd(date)) %in% c(1,7), "weekend", "weekday"))) %>% select(steps, date, interval, DayClass)
```

Now let's create a summary with the mean steps, grouped by weekday/weekend and interval.

```{r}
wdayIntervalSum <- activitynoNA %>% group_by(DayClass, interval) %>% summarise(mean_steps = mean(steps))
wdayIntervalSum
```

Now let's create a panel plot using the summary tibble and ggplot2.

```{r}
g <- ggplot(data = wdayIntervalSum, aes(interval, mean_steps))
g + geom_line(color = "navy blue") + facet_grid(DayClass~.) + ylab("Average Steps") + ggtitle("Average steps per interval on weekdays & weekends") + theme_bw()
```
